---
title: "The anievent data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
```

## What is an `anievent`?

`anievent` is the sibling of [`aniframe`](aniframe-structure.html) for
**behavioural events**: sleep bouts, vocalisations, manual scoring from
BORIS, the output of segmentation models. Where an `aniframe` records
position frame-by-frame, an `anievent` records intervals
(`start` → `stop`) or instants (`start == stop`) of labelled behaviour.

The class is a true sibling — it does **not** inherit from `aniframe`,
but it shares the metadata substrate, so `get_metadata()`,
`set_metadata()`, and the rest of the metadata API work on both.

## Channels

A **channel** is a single mutually-exclusive categorical track of
behaviour. Within one channel, a subject can carry only one label at
a time: a sleep-state channel is either `REM` or `wake`, never both
simultaneously. If two behaviours **can** co-occur — say grooming
while walking — they belong on **separate channels**, not as two
values of the same channel. That mutual-exclusivity constraint is
what makes a channel a channel.

This means how you design your channels is a modelling choice. BORIS
calls them "behavioural categories" in the ethogram; segmentation
pipelines call them "label streams"; we call them channels in this
package.

## Anatomy at a glance

```{r}
ae <- anievent(
  individual = c(1L, 1L, 1L, 2L),
  channel = c("behaviour", "behaviour", "call", "behaviour"),
  value = c("REM", "wake", "alarm", "REM"),
  start = c(3, 14, 4.5, 1),
  stop = c(9, 19, 4.5, 6)
)
ae
```

Two things to spot:

1. The **header rows** prefixed with `#` come from the metadata
   attribute — identity columns and the event channels carried by the
   `channel` column.
2. The **data columns** are the long-format event records: one row per
   bout (or instant), with the channel name in `channel` and the
   label in `value`.

## Required schema

An `anievent` has four mandatory columns plus identity columns that
travel via `variables_what`:

| Column       | Type      | Required? | Role                                       |
|--------------|-----------|-----------|--------------------------------------------|
| `channel`    | character | yes       | which event channel                        |
| `value`      | factor    | yes       | the label                                  |
| `start`      | numeric   | yes       | bout start (or instant)                    |
| `stop`       | numeric   | yes       | bout stop (equals `start` for point events) |
| identity     | as source | optional  | e.g. `individual`, `track`, `subject`      |
| `modifiers`  | list-column of `character` | optional | flat per-event modifier values |

Identity columns are auto-detected from a known list
(`model`, `individual`, `track`, `subject`), or you can supply any
name via `variables_what`. An `anievent` with no identity column at
all is fine for single-subject experiments.

### Multiple observations / trials / sessions

Behavioural coding often spans several clips or trials, each with its
own time origin (`start = 0` means "the start of *this* clip", not "the
start of the experiment"). To carry that context, add an `observation`,
`session`, or `trial` column — those three names are auto-detected
into `variables_when` alongside `start` and `stop`, and the column is
coerced to factor / integer the same way aniframe does it:

```{r}
ae_multi <- anievent(
  individual = c(1L, 1L, 1L, 1L),
  observation = c("clip_a", "clip_a", "clip_b", "clip_b"),
  channel =c("behaviour", "behaviour", "behaviour", "behaviour"),
  value = c("REM", "wake", "REM", "wake"),
  start = c(3, 14, 1, 7),
  stop = c(9, 19, 5, 12)
)
get_metadata(ae_multi, "variables_when")
ae_multi
```

The `start`/`stop` values are local to each observation — both clips
have an event starting near the beginning, but those are separate
moments in real time. Anything downstream that needs to combine clips
onto a shared timeline (plots, cross-clip metrics) will need an
observation-level offset table; the class itself just carries the key.

### Time unit

`unit_time` is carried in the shared metadata substrate, so an
`anievent` declares the unit of its `start`/`stop` values just like
an `aniframe` declares the unit of its `time` column (`"frame"`,
`"ms"`, `"s"`, `"m"`, `"h"`, …). `set_unit_time()` and
`set_sampling_rate()` dispatch on both classes — on an anievent they
scale `start` and `stop` instead of `time`:

```{r}
ae_frames <- anievent(
  individual = 1L,
  channel = c("behaviour", "behaviour"),
  value = c("REM", "wake"),
  start = c(30, 150),
  stop = c(60, 300)
)
ae_seconds <- set_sampling_rate(ae_frames, sampling_rate = 30)
ae_seconds
```

## State vs point events

Two temporal kinds:

- **State events** — interval-valued, durative (`stop > start`). A
  sleep bout, a grooming bout.
- **Point events** — instantaneous (`stop == start`). A startle, a
  vocalisation onset.

The `anievent` class is intentionally **type-agnostic**: it does not
require you to declare which channels are state and which are point.
The `channel` column already names the channels, and the duration of
each row tells you everything the class needs.

Type only becomes load-bearing at conversion / plotting / metric time:

- when converting an `anievent` to a per-frame `aniframe` (so the
  conversion knows whether to fill the whole interval or just a single
  frame),
- when plotting (state → coloured band; point → marker),
- when computing metrics that differ by type.

Those operations live in companion packages (`animetric`, `anivis`)
and will accept type declarations at call time. The metadata field
`variables_event = list(state = ..., point = ...)` exists on the
shared metadata substrate for callers that want to record the
distinction explicitly, but the `anievent` class itself does not
enforce it.

## Modifiers

A **modifier** is a per-event attribute set at coding time — in BORIS,
the values chosen from a behaviour's modifier sets. A grooming bout
might carry the values `"limb"` and `"whisker"`; an alarm call might
carry `"high"`.

Modifiers are stored as an optional `modifiers` list-column, where
each cell is a **flat character vector** (length zero, one, or many).
This matches how BORIS exports modifier data: the set-to-value
structure from the ethogram is collapsed into a single bag of values
per event, so a modifier cell is just the values picked, with no
set-key wrapping.

```{r}
ae_with_mods <- anievent(
  individual = c(1L, 1L, 1L),
  channel =c("behaviour", "behaviour", "call"),
  value = c("REM", "REM", "alarm"),
  start = c(3, 14, 4.5),
  stop = c(9, 19, 4.5),
  modifiers = list(
    c("limb", "whisker"),
    "tail",
    character()
  )
)
ae_with_mods
```

Standard dplyr verbs round-trip the list-column without further
ceremony:

```{r}
ae_with_mods |>
  dplyr::filter(channel == "behaviour")
```

A helper to widen this list-column into one logical column per
distinct modifier value (so filtering or faceting on a single
modifier value is straightforward) is planned but not yet
implemented.

## Validation

`validate_anievent()` re-checks the structural invariants on demand,
returning the input invisibly on success and erroring otherwise:

```{r}
#| error: true
# Happy path: returns invisibly, no output
validate_anievent(ae_with_mods)

# Negative interval is caught
bad <- ae_with_mods
bad$stop[1] <- bad$start[1] - 1
validate_anievent(bad)
```

The validator confirms required columns are present with the right
types, that `stop >= start` for every row, and that any `modifiers`
column is well-formed. It deliberately does **not** check
type-linked invariants (e.g. point rows having `start == stop`) —
those belong to the conversion code that knows which channels are
typed as what.

## Conversions to and from `aniframe`

The two structures are pivot-duals of each other. An `aniframe`'s
event columns (declared in `variables_event`) hold one label per
frame, wide; an `anievent` holds one record per bout, long. Two
verbs go between them.

### `as_anievent()` — `aniframe` → `anievent`

For each column listed in `variables_event$state`, contiguous runs of
the same label within each `(individual, observation, …)` group are
run-length-encoded into bouts. For each column in `variables_event$point`,
every non-`NA` frame becomes a single row with `start == stop`.

```{r}
af <- aniframe(
  individual = rep(1L, 8),
  time = 1:8,
  x = 1:8,
  y = 1:8,
  behaviour = factor(
    c("REM", "REM", "REM", "wake", "wake", "REM", "REM", NA),
    levels = c("REM", "wake")
  )
)
af <- set_metadata(
  af,
  variables_event = list(state = "behaviour", point = character())
)
as_anievent(af)
```

The resulting anievent inherits `unit_time` and `sampling_rate` from
the host. `stop` is the time of the last frame of each bout (inclusive,
frame-aligned).

### `add_events()` — `anievent` → `aniframe` (host-first)

Joins an `anievent` onto an `aniframe`'s per-frame grid, adding one
column per channel. For each row of the host, the new column carries
the `value` of the active bout for the matching subject (and
observation, etc.); `NA` outside any bout.

```{r}
af <- aniframe(individual = 1L, time = 1:10, x = 1:10, y = 1:10)
ae <- anievent(
  individual = 1L,
  channel = c("behaviour", "behaviour", "call"),
  value = c("REM", "wake", "alarm"),
  start = c(1, 5, 3),
  stop = c(4, 10, 3)
)
add_events(af, ae)
```

State vs point is auto-detected from the bouts (a channel is `point`
iff every bout has `start == stop`). The resulting aniframe's
`variables_event` is updated accordingly. The auto-detect rule can be
overridden — pass `variables_event = list(state = ..., point = ...)`
to `add_events()` or set it on the `events`' own metadata. This is
important when a state channel happens to have only single-frame
bouts (e.g. flapping VAME motifs) and would otherwise be misread as
point.

Unit reconciliation happens transparently when host and events differ:
SI ↔ SI is auto-converted; crossing the `"frame"` ↔ SI boundary needs
a `sampling_rate` on either side (errors otherwise).

### Splitting and bundling

An `anievent` can have overlapping bouts on the same channel — that's
allowed at the long-form layer. An `aniframe`'s factor column can't
hold two values at the same frame, so `add_events()` greedy-splits
overlapping bouts into numbered sub-columns: track 1 stays as
`<channel>`, track 2 becomes `<channel>_2`, and so on. Each output
column is itself mutually exclusive.

`as_anievent.aniframe()` reverses the split — when both `<channel>`
and `<channel>_<n>` columns are declared in `variables_event`, they're
bundled back under one logical channel on the resulting anievent. So
the round-trip preserves the original channel grouping, even though
the dense form temporarily uses multiple columns.

### Modifiers round-trip

If the anievent carries a `modifiers` list-column, `add_events()`
broadcasts each bout's modifier vector across its frames as a
parallel `<channel>_modifiers` list-column on the host. Channels
without any non-empty modifier cells don't add a column.

`as_anievent()` reverses the operation: a `<channel>_modifiers`
column on the source aniframe is gathered back into the resulting
anievent's `modifiers` cells, one per row.

```{r}
ae_with_mods <- anievent(
  individual = c(1L, 1L),
  channel = c("behaviour", "behaviour"),
  value = c("REM", "wake"),
  start = c(1, 5),
  stop = c(4, 10),
  modifiers = list(c("limb", "whisker"), "tail")
)
af_with_mods <- add_events(
  aniframe(individual = 1L, time = 1:10, x = 1:10, y = 1:10),
  ae_with_mods
)
af_with_mods |> dplyr::select(time, behaviour, behaviour_modifiers)
```

## Where things live in the ecosystem

| Concern                                 | Planned home | Status     |
|-----------------------------------------|--------------|------------|
| The `anievent` class                    | `aniframe`   | ✅ implemented |
| Conversions to / from `aniframe`        | `aniframe`   | ✅ implemented |
| Readers (BORIS, Solomon, …)             | `aniread`    | planned    |
| Bout-summary metrics                    | `animetric`  | planned    |
| Plots (state band, point marker)        | `anivis`     | planned    | -->
