---
title: "The anievent data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
```

## What is an `anievent`?

`anievent` is the sibling of [`aniframe`](aniframe-structure.html) for
**behavioural events**: sleep bouts, vocalisations, manual scoring from
BORIS, the output of segmentation models. Where an `aniframe` records
position frame-by-frame, an `anievent` records intervals
(`start` → `stop`) or instants (`start == stop`) of labelled behaviour.

The class is a true sibling — it does **not** inherit from `aniframe`,
but it shares the metadata substrate, so `get_metadata()`,
`set_metadata()`, and the rest of the metadata API work on both.

## Anatomy at a glance

```{r}
ae <- anievent(
  individual = c(1L, 1L, 1L, 2L),
  variable = c("behaviour", "behaviour", "call", "behaviour"),
  value = c("REM", "wake", "alarm", "REM"),
  start = c(3, 14, 4.5, 1),
  stop = c(9, 19, 4.5, 6)
)
ae
```

Two things to spot:

1. The **header rows** prefixed with `#` come from the metadata
   attribute — identity columns and the event channels carried by the
   `variable` column.
2. The **data columns** are the long-format event records: one row per
   bout (or instant), with the channel name in `variable` and the
   label in `value`.

## Required schema

An `anievent` has four mandatory columns plus identity columns that
travel via `variables_what`:

| Column       | Type      | Required? | Role                                       |
|--------------|-----------|-----------|--------------------------------------------|
| `variable`   | character | yes       | which event channel                        |
| `value`      | factor    | yes       | the label                                  |
| `start`      | numeric   | yes       | bout start (or instant)                    |
| `stop`       | numeric   | yes       | bout stop (equals `start` for point events) |
| identity     | as source | optional  | e.g. `individual`, `track`, `subject`      |
| `modifiers`  | list-column of `character` | optional | flat per-event modifier values |

Identity columns are auto-detected from a known list
(`model`, `individual`, `track`, `subject`), or you can supply any
name via `variables_what`. An `anievent` with no identity column at
all is fine for single-subject experiments.

### Multiple observations / trials / sessions

Behavioural coding often spans several clips or trials, each with its
own time origin (`start = 0` means "the start of *this* clip", not "the
start of the experiment"). To carry that context, add an `observation`,
`session`, or `trial` column — those three names are auto-detected
into `variables_when` alongside `start` and `stop`, and the column is
coerced to factor / integer the same way aniframe does it:

```{r}
ae_multi <- anievent(
  individual = c(1L, 1L, 1L, 1L),
  observation = c("clip_a", "clip_a", "clip_b", "clip_b"),
  variable = c("behaviour", "behaviour", "behaviour", "behaviour"),
  value = c("REM", "wake", "REM", "wake"),
  start = c(3, 14, 1, 7),
  stop = c(9, 19, 5, 12)
)
get_metadata(ae_multi, "variables_when")
ae_multi
```

The `start`/`stop` values are local to each observation — both clips
have an event starting near the beginning, but those are separate
moments in real time. Anything downstream that needs to combine clips
onto a shared timeline (plots, cross-clip metrics) will need an
observation-level offset table; the class itself just carries the key.

## State vs point events

Two temporal kinds:

- **State events** — interval-valued, durative (`stop > start`). A
  sleep bout, a grooming bout.
- **Point events** — instantaneous (`stop == start`). A startle, a
  vocalisation onset.

The `anievent` class is intentionally **type-agnostic**: it does not
require you to declare which channels are state and which are point.
The `variable` column already names the channels, and the duration of
each row tells you everything the class needs.

Type only becomes load-bearing at conversion / plotting / metric time:

- when converting an `anievent` to a per-frame `aniframe` (so the
  conversion knows whether to fill the whole interval or just a single
  frame),
- when plotting (state → coloured band; point → marker),
- when computing metrics that differ by type.

Those operations live in companion packages (`animetric`, `anivis`)
and will accept type declarations at call time. The metadata field
`variables_event = list(state = ..., point = ...)` exists on the
shared metadata substrate for callers that want to record the
distinction explicitly, but the `anievent` class itself does not
enforce it.

## Modifiers

A **modifier** is a per-event attribute set at coding time — in BORIS,
the values chosen from a behaviour's modifier sets. A grooming bout
might carry the values `"limb"` and `"whisker"`; an alarm call might
carry `"high"`.

Modifiers are stored as an optional `modifiers` list-column, where
each cell is a **flat character vector** (length zero, one, or many).
This matches how BORIS exports modifier data: the set-to-value
structure from the ethogram is collapsed into a single bag of values
per event, so a modifier cell is just the values picked, with no
set-key wrapping.

```{r}
ae_with_mods <- anievent(
  individual = c(1L, 1L, 1L),
  variable = c("behaviour", "behaviour", "call"),
  value = c("REM", "REM", "alarm"),
  start = c(3, 14, 4.5),
  stop = c(9, 19, 4.5),
  modifiers = list(
    c("limb", "whisker"),
    "tail",
    character()
  )
)
ae_with_mods
```

Standard dplyr verbs round-trip the list-column without further
ceremony:

```{r}
ae_with_mods |>
  dplyr::filter(variable == "behaviour")
```

A helper to widen this list-column into one logical column per
distinct modifier value (so filtering or faceting on a single
modifier value is straightforward) is planned but not yet
implemented.

## Validation

`validate_anievent()` re-checks the structural invariants on demand,
returning the input invisibly on success and erroring otherwise:

```{r}
#| error: true
# Happy path: returns invisibly, no output
validate_anievent(ae_with_mods)

# Negative interval is caught
bad <- ae_with_mods
bad$stop[1] <- bad$start[1] - 1
validate_anievent(bad)
```

The validator confirms required columns are present with the right
types, that `stop >= start` for every row, and that any `modifiers`
column is well-formed. It deliberately does **not** check
type-linked invariants (e.g. point rows having `start == stop`) —
those belong to the conversion code that knows which channels are
typed as what.

## Relationship to `aniframe`

The two structures are pivot-duals of each other. An `aniframe`'s
event columns (declared in `variables_event`) hold one label per
frame, wide; an `anievent` holds one record per bout, long.

::: callout-note
## Planned, not yet available

Conversions between the two are the natural next step:

- run-length-encoding the wide per-frame event columns of an
  `aniframe` into the long bout form of an `anievent`,
- broadcasting an `anievent`'s bouts back onto an `aniframe`'s
  per-frame grid.

Both directions will live in `aniframe`, since the package owns both
classes. A BORIS reader (producing an `anievent` directly) is planned
for `aniread`. None of these are implemented yet.
:::

## Where things live in the ecosystem

Only the `anievent` class itself ships today; everything else is
planned for follow-up work in the respective packages.

| Concern                                 | Planned home | Status     |
|-----------------------------------------|--------------|------------|
| The `anievent` class                    | `aniframe`   | ✅ implemented |
| Conversions to / from `aniframe`        | `aniframe`   | planned    |
| Readers (BORIS, Solomon, …)             | `aniread`    | planned    |
| Bout-summary metrics                    | `animetric`  | planned    |
| Plots (state band, point marker)        | `anivis`     | planned    |
