---
title: "The anievent data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
```

## What is an `anievent`?

`anievent` is the sibling of [`aniframe`](aniframe-structure.html) for
**behavioural events**: sleep bouts, vocalisations, manual scoring from
BORIS, the output of segmentation models. Where an `aniframe` records
position frame-by-frame, an `anievent` records intervals
(`start` → `stop`) or instants (`start == stop`) of labelled behaviour.

The class is a true sibling — it does **not** inherit from `aniframe`,
but it shares the metadata substrate, so `get_metadata()`,
`set_metadata()`, and the rest of the metadata API work on both.

## Anatomy at a glance

```{r}
ae <- anievent(
  individual = c(1L, 1L, 1L, 2L),
  variable = c("behaviour", "behaviour", "call", "behaviour"),
  value = c("REM", "wake", "alarm", "REM"),
  start = c(3, 14, 4.5, 1),
  stop = c(9, 19, 4.5, 6)
)
ae
```

Two things to spot:

1. The **header rows** prefixed with `#` come from the metadata
   attribute — identity columns and the event channels carried by the
   `variable` column.
2. The **data columns** are the long-format event records: one row per
   bout (or instant), with the channel name in `variable` and the
   label in `value`.

## Required schema

An `anievent` has four mandatory columns plus identity columns that
travel via `variables_what`:

| Column        | Type      | Role                                         |
|---------------|-----------|----------------------------------------------|
| `variable`    | character | which event channel / column                 |
| `value`       | factor    | the label                                    |
| `start`       | numeric   | bout start (or instant)                      |
| `stop`        | numeric   | bout stop (equals `start` for point events)  |
| `individual`* | as source | identity (default; rename via `variables_what`) |
| `modifiers`*  | list      | optional per-event values (BORIS-style)      |

\* optional.

Identity columns are auto-detected from a known list
(`model`, `individual`, `track`, `subject`), or you can supply any
name via `variables_what`. An `anievent` with no identity column at
all is fine for single-subject experiments.

## State vs point events

Two temporal kinds:

- **State events** — interval-valued, durative (`stop > start`). A
  sleep bout, a grooming bout.
- **Point events** — instantaneous (`stop == start`). A startle, a
  vocalisation onset.

The `anievent` class is intentionally **type-agnostic**: it does not
require you to declare which channels are state and which are point.
The `variable` column already names the channels, and the duration of
each row tells you everything the class needs.

Type only becomes load-bearing at conversion / plotting / metric time:

- when converting an `anievent` to a per-frame `aniframe` (so the
  conversion knows whether to fill the whole interval or just a single
  frame),
- when plotting (state → coloured band; point → marker),
- when computing metrics that differ by type.

Those operations live in companion packages (`animetric`, `anivis`)
and will accept type declarations at call time. The metadata field
`variables_event = list(state = ..., point = ...)` exists on the
shared metadata substrate for callers that want to record the
distinction explicitly, but the `anievent` class itself does not
enforce it.

## Modifiers

A **modifier** is a per-event attribute set at coding time — in BORIS,
the values chosen from a behaviour's modifier sets. Examples:

- `parts_moved = c("limb", "whisker", "tail")` on a movement bout
- `intensity = "high"` on an alarm call

Modifiers are stored as an optional `modifiers` list-column, where
each cell is a **character vector** (length zero, one, or many). This
matches the BORIS export format: one event row carries a flat list
of selected modifier values.

```{r}
ae_with_mods <- anievent(
  individual = c(1L, 1L, 1L),
  variable = c("behaviour", "behaviour", "call"),
  value = c("REM", "REM", "alarm"),
  start = c(3, 14, 4.5),
  stop = c(9, 19, 4.5),
  modifiers = list(
    c("limb", "whisker"),
    "tail",
    character()
  )
)
ae_with_mods
```

Standard dplyr verbs round-trip the list-column without further
ceremony:

```{r}
ae_with_mods |>
  dplyr::filter(variable == "behaviour")
```

A future `expand_modifiers()` helper (next slice) will turn this
list-column into one-hot logical columns for analysis.

## Validation

`validate_anievent()` re-checks the structural invariants on demand:

```{r}
validate_anievent(ae_with_mods)
```

The validator confirms required columns are present with the right
types, that `stop >= start` for every row, and that any `modifiers`
column is well-formed. It deliberately does **not** check
type-linked invariants (e.g. point rows having `start == stop`) —
those belong to the conversion code that knows which channels are
typed as what.

## Relationship to `aniframe`

The two structures are pivot-duals of each other. An `aniframe`'s
event columns (declared in `variables_event`) hold one label per
frame, wide; an `anievent` holds one record per bout, long.

Conversions between the two are the natural next step:

- `as_anievent(aniframe)` will run-length-encode the wide event
  columns into long bouts.
- `add_events(aniframe, anievent)` will broadcast bouts back onto a
  per-frame grid.

Both verbs live in `aniframe` (since they own both classes); they
will land in the next slice along with the BORIS reader (in
`aniread`).

## Where things live in the ecosystem

| Concern                              | Package      |
|--------------------------------------|--------------|
| The `anievent` class + conversions   | `aniframe`   |
| Readers (BORIS, Solomon, …)          | `aniread`    |
| Bout-summary metrics                 | `animetric`  |
| Plots (state band, point marker)     | `anivis`     |
| Segmentation models that produce events | `anisegment` |
