---
title: "The aniframe data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
library(dplyr)
```

## Overview

`aniframe` is the in-memory representation of *tidy movement data* — an
attempt to give the heterogeneous world of movement-tracking outputs a
single shape that downstream tools can rely on. The conceptual case for
why this shape and not another is laid out in an upcoming paper; this
page is the practical reference for what an aniframe actually contains
and how the pieces fit together.

The page has two halves:

- **Part I — Structure** describes the *data*: which columns belong in
  an aniframe, how they answer the questions *what is moving*, *when*,
  and *where*, and how the spatial columns determine the coordinate
  system.
- **Part II — Metadata** describes the *attribute* attached alongside
  the data — sampling rate, units, origin, frame height, connections —
  and the functions that read and update each field.

If you only want to *use* an aniframe (filter, mutate, plot), you mostly
don't need this page — the class plays nicely with dplyr and base R.
This page is for when you want to know what's actually in there.

# Part I — Structure

## Anatomy at a glance

The cheapest way to see an aniframe is to print one.

```{r}
data <- example_aniframe(
  n_obs = 5,
  n_individuals = 2,
  n_keypoints = 3,
  n_dims = 2
)
data
```

Three things to spot:

1. The **header rows** prefixed with `#` — these come from the metadata
   attribute and summarise what the object contains (individuals,
   keypoints, recording duration when known, sampling rate, etc.).
2. The **data columns** — a tidy table where every row records one
   entity at one timepoint at one position.
3. The class line at the bottom — an aniframe is a `tibble` underneath
   and works with `dplyr`, `ggplot2`, base subsetting, etc. without
   ceremony.

The rest of Part I unpacks the columns. Part II unpacks the header.

## The three slots: what / when / where

Every row of an aniframe answers three questions:

| Question | Slot          | Resolves   | Default columns    |
|----------|---------------|------------|--------------------|
| **What** is moving? | Identity (`variables_what`) | An entity   | `individual`, `keypoint` |
| **When**?           | Temporal (`variables_when`) | A timepoint | `time`                   |
| **Where**?          | Spatial (`variables_where`) | A position  | `x`, `y`                 |

The combination of all identity and temporal columns forms a composite
key that uniquely identifies each row; the spatial columns are what
that row *records* about that entity at that timepoint.

Aniframe stores which columns play which role in the metadata, so
downstream code can introspect:

```{r}
md <- get_metadata(data)
md$variables_what
md$variables_when
md$variables_where
```

### Customising the slots

Most readers populate these from the source data, but you can override
them when constructing manually. Anything in
`c("model", "individual", "track", "keypoint")` is auto-detected as
identity; anything in `c("session", "trial", "time")` as temporal;
spatial columns are detected from the column names below.

```{r}
df <- data.frame(
  track = rep(c("A", "B", "C"), each = 4),
  trial = rep(1:2, each = 6),
  time = rep(1:2, 6),
  x = runif(12),
  y = runif(12)
)

custom <- as_aniframe(
  df,
  variables_what = "track",
  variables_when = c("trial", "time")
)
custom
```

`time` is the only temporal column with a special status: it must
always be present, and downstream operations (smoothing, derivatives)
assume it's monotonically ordered within each entity.

## Coordinate systems

The set of spatial columns determines the coordinate system. Aniframe
recognises four families:

::: {.panel-tabset}

### Cartesian

`x`, `y`, and optionally `z`. The number of columns picks the
dimensionality (`cartesian_1d`, `cartesian_2d`, `cartesian_3d`).

```{r}
cart <- aniframe(
  individual = 1L, time = 1:3,
  x = c(0, 1, 2), y = c(0, 1, 4), z = c(0, 0, 1)
)
get_metadata(cart, "coordinate_system")
```

### Polar

`rho` (radius) and `phi` (angle). 2D only.

```{r}
pol <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi)
)
get_metadata(pol, "coordinate_system")
```

### Cylindrical

`rho`, `phi`, and `z`. The presence of `z` alongside `rho`+`phi`
distinguishes it from polar.

```{r}
cyl <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi), z = c(0, 1, 2)
)
get_metadata(cyl, "coordinate_system")
```

### Spherical

`rho`, `phi`, and `theta`.

```{r}
sph <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi), theta = c(0, pi / 4, pi / 2)
)
get_metadata(sph, "coordinate_system")
```

:::

Detection runs polar-family first (`rho` + `phi` together is the
signature), with Cartesian as the fallback. This means a column named
`z` doesn't accidentally make cylindrical data look 1D-Cartesian.

For programmatic checks, every coordinate system has both an
`is_<system>()` predicate (returns logical) and an
`ensure_is_<system>()` guard (errors otherwise):

```{r}
is_polar(pol)
is_cartesian_2d(cart)
```

# Part II — Metadata

## The metadata attribute

Every aniframe carries a metadata list as an R attribute. You can see
the full thing by printing it directly:

```{r}
get_metadata(data)
```

The fields and their defaults are defined in a single place,
`default_metadata()` — that's the canonical source of truth for what an
aniframe's metadata looks like.

```{r}
str(default_metadata(), max.level = 1)
```

The fields fall into a few groups:

| Group              | Fields                                                              |
|--------------------|---------------------------------------------------------------------|
| **Provenance**     | `source`, `source_version`, `filename`, `start_datetime`            |
| **Sampling**       | `sampling_rate`                                                     |
| **Units**          | `unit_space`, `unit_time`, `unit_angle`                             |
| **Frame of reference** | `reference_frame`, `coordinate_system`, `origin`, `y_height`    |
| **Slot vocabulary**| `variables_what`, `variables_when`, `variables_where`               |
| **Relationships**  | `connections`                                                       |

`filename` accepts a character vector — readers like
`aniread::read_trackball()` populate it with both source paths.

## Reading and writing metadata

`get_metadata()` and `set_metadata()` are the workhorses.

```{r}
get_metadata(data, "sampling_rate")

data <- set_metadata(data, sampling_rate = 30, source = "deeplabcut")
get_metadata(data, "sampling_rate")
get_metadata(data, "source")
```

`set_metadata()` validates the input — factor fields are checked
against their permitted levels, and unknown fields are rejected.

For fields with non-trivial side effects, prefer the dedicated setters
listed below. They keep related fields consistent and (where it makes
sense) update the data columns to match.

| Setter                | Touches                                                        |
|-----------------------|----------------------------------------------------------------|
| `set_unit_space()`    | converts `x`/`y`/`z` between length units                      |
| `set_unit_time()`     | converts `time` between time units                             |
| `set_unit_angle()`    | converts `phi`/`theta` (auto) and any extra `cols` you supply  |
| `set_sampling_rate()` | flips `unit_time` from frames to seconds and rescales `time`   |
| `set_origin()`        | flips the y-axis around `y_height` when changing convention    |
| `set_y_height()`      | sets the recorded frame height used by `set_origin()`          |
| `set_connections()`   | replaces the connection table for a variable (see below)       |

## Coordinate origin

For 2D image-derived data there's an annoying convention split: most
image / video tooling uses the **top-left** corner as `(0, 0)` (y
increases downward), while plotting and most maths uses the
**bottom-left** corner (y increases upward). Aniframe records which one
your data uses in the `origin` field, with permitted values
`c("bottom_left", "top_left")`.

`set_origin()` does the actual flip when you change convention. It
needs to know the frame height to compute `y_new = y_height - y_old`,
so `y_height` must be set first. Readers populate it automatically; for
manually-constructed aniframes, `as_aniframe()` falls back to `max(y)`
and `set_y_height()` lets you override that with the true value.

```{r}
img <- aniframe(
  individual = 1L, time = 1:4,
  x = c(0, 10, 20, 30),
  y = c(50, 100, 150, 200)
) |>
  set_y_height(1080)

img$y
img <- set_origin(img, "top_left")
img$y # reflected: 1080 - original y
```

## Units

Spatial, temporal, and angular units each have their own setter.
Conversions between standard units are automatic; conversions from
"unknown" / `frame` / `px` units require an explicit calibration
factor (or, for time, a `sampling_rate`).

```{r}
data <- example_aniframe(n_dims = 2) |>
  set_metadata(unit_space = "mm")

data_cm <- set_unit_space(data, to_unit = "cm")
get_metadata(data_cm, "unit_space")
```

```{r}
data <- example_aniframe() # default unit_time = "frame"
data_s <- set_sampling_rate(data, sampling_rate = 30)
get_metadata(data_s, "unit_time") # now "s"
range(data_s$time) # frames divided by fps
```

Spatial angular columns (`phi`, `theta`) are converted automatically by
`set_unit_angle()` whenever they're present. Pass `cols` only for
non-spatial angular columns (e.g. heading direction).

```{r}
pol <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi)
)
pol$phi
pol_deg <- set_unit_angle(pol, to_unit = "deg")
pol_deg$phi
```

## Connections (skeletons & networks)

`metadata$connections` records the relationships *between* values of an
identity or temporal variable — most commonly a skeleton (edges between
keypoints), but it could also be a social network (edges between
individuals) or a session graph.

The storage shape is a named list keyed by the variable, with each
entry being a 2-column `from`/`to` tibble:

```{r}
fish <- example_aniframe(n_keypoints = 7) |>
  set_connections(
    list(
      c("head", "neck"),
      c("neck", "shoulder_right"),
      c("neck", "shoulder_left"),
      c("shoulder_right", "abdomen"),
      c("shoulder_left", "abdomen")
    )
  )

get_connections(fish, "keypoint")
```

The order is preserved as supplied, so downstream consumers can treat
the table as either directed (parent → child kinematic chains) or
undirected (skeleton edges) depending on what they need.

Inputs come in three flavours — pick whichever is most natural:

```{r}
#| eval: false
# Implicit position
set_connections(data, list(c("head", "neck")))

# Explicit names within each pair
set_connections(data, list(c(from = "head", to = "neck")))

# 2-column data.frame
set_connections(data, data.frame(from = "head", to = "neck"))
```

For incremental editing, `add_connections()` and `remove_connections()`
take vectors so multiple edges go in or out in one call:

```{r}
fish <- add_connections(
  fish,
  from = c("abdomen", "abdomen"),
  to   = c("hip_right", "hip_left")
)
nrow(get_connections(fish, "keypoint"))
```

When a `from`/`to` value isn't found in the corresponding column,
`set_connections()` and `add_connections()` emit a warning (typo
catcher) but keep the connection — the value may legitimately be
missing in this particular recording.

## Persistence

The aniframe ecosystem uses Apache Parquet as its on-disk format. The
data table is stored as a Parquet file in the obvious way; the
metadata list (including the nested `connections` field) is serialised
into the Parquet schema metadata using R's native serialisation, so the
in-memory structure round-trips losslessly.

```{r}
#| eval: false
library(aniread)

write_aniframe(fish, "fish.parquet")
restored <- read_aniframe("fish.parquet")
identical(get_metadata(fish), get_metadata(restored))
#> [1] TRUE
```

The wire-format details (Parquet schema metadata key, column types,
how each metadata field is encoded) are documented by the `aniread`
package, which owns the readers and writers.

---

## Reference appendix

For sister-package authors, the contract is:

- **Class**: an aniframe is a `tibble` with the additional class
  `"aniframe"`. Use `is_aniframe()` to test, `ensure_is_aniframe()` to
  guard.
- **Required column**: `time`. Identity / spatial columns must be
  present in some recognised form — see `as_aniframe()`.
- **Metadata location**: `attr(data, "metadata")`, structured as an
  `aniframe_metadata` list.
- **Canonical metadata schema**: `default_metadata()`. Field names,
  types, factor levels, and defaults are all defined there in one
  place. New aniframe versions extend this list rather than mutate it
  in place.
- **Validation**: `ensure_valid_metadata()` checks that all fields are
  present and have the right class.
