---
title: "The aniframe data structure"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
```

## What is an `aniframe`?

The _aniframe_ package defines `aniframe` — the foundational data
structure of the _animovement_ ecosystem. Every sister package
(_aniread_ for I/O, _animetric_ for kinematics, …) is built around
consuming and producing `aniframe` objects, so the shape that an
`aniframe` takes is what holds the ecosystem together.

The shape itself follows the principles of *tidy movement data* (laid
out in an upcoming paper). The short version: every row records one
entity at one timepoint at one position, and the columns split into
three semantic slots that answer **what** is moving, **when**, and
**where**.

This article covers the data shape. Companion articles cover the
[metadata attribute](aniframe-metadata.html) and the
[connections field](aniframe-connections.html) for skeletons and
networks.

## Anatomy at a glance

The cheapest way to see an `aniframe` is to print one.

```{r}
data <- example_aniframe(
  n_obs = 5,
  n_individuals = 2,
  n_keypoints = 3,
  n_dims = 2
)
data
```

Two things to spot:

1. The **header rows** prefixed with `#` come from the metadata
   attribute and summarise what the object contains — individuals,
   keypoints, recording duration when known, sampling rate, etc.
2. The **data columns** form a tidy table where every row records one
   entity at one timepoint at one position.

`aniframe` inherits from `tbl_df` and `data.frame`, so dplyr verbs,
ggplot2, and base subsetting all work on it directly.

The rest of this article unpacks the columns.

## The three slots: what / when / where

Every row of an `aniframe` answers three questions:

| Question | Slot          | Resolves   | Default columns          |
|----------|---------------|------------|--------------------------|
| **What** is moving? | Identity (`variables_what`) | An entity   | `individual`, `keypoint` |
| **When**?           | Temporal (`variables_when`) | A timepoint | `time`                   |
| **Where**?          | Spatial (`variables_where`) | A position  | `x`, `y`                 |

The combination of all identity and temporal columns forms a composite
key that uniquely identifies each row; the spatial columns are what
that row *records* about that entity at that timepoint.

`aniframe` stores which columns play which role in the metadata, so
downstream code can introspect:

```{r}
md <- get_metadata(data)
md$variables_what
md$variables_when
md$variables_where
```

### Customising the slots

Most readers populate these from the source data, but you can override
them when constructing manually. By default, `as_aniframe()` recognises
`c("model", "individual", "track", "keypoint")` as identity columns and
`c("session", "trial", "time")` as temporal columns. Pass the slot
arguments explicitly to use other names:

```{r}
df <- data.frame(
  track = rep(c("A", "B", "C"), each = 4),
  trial = rep(1:2, each = 6),
  time = rep(1:2, 6),
  x = runif(12),
  y = runif(12)
)

custom <- as_aniframe(
  df,
  variables_what = "track",
  variables_when = c("trial", "time")
)
custom
```

`time` is the only temporal column with a special status: it must
always be present, and downstream operations (smoothing, derivatives)
assume it's monotonically ordered within each entity.

## Coordinate systems

The set of spatial columns determines the coordinate system. `aniframe`
recognises four families:

| System       | Columns                          |
|--------------|----------------------------------|
| Cartesian    | `x`, `y`, optionally `z`         |
| Polar        | `rho`, `phi`                     |
| Cylindrical  | `rho`, `phi`, `z`                |
| Spherical    | `rho`, `phi`, `theta`            |

The number of Cartesian columns picks the dimensionality
(`cartesian_1d`, `cartesian_2d`, `cartesian_3d`):

```{r}
cart <- aniframe(
  individual = 1L, time = 1:3,
  x = c(0, 1, 2), y = c(0, 1, 4), z = c(0, 0, 1)
)
get_metadata(cart, "coordinate_system")
```

`rho` + `phi` indicates a polar-family system; the third column (if
any) distinguishes the variant:

```{r}
pol <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi)
)
get_metadata(pol, "coordinate_system")

cyl <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi), z = c(0, 1, 2)
)
get_metadata(cyl, "coordinate_system")

sph <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi), theta = c(0, pi / 4, pi / 2)
)
get_metadata(sph, "coordinate_system")
```

For programmatic checks, every coordinate system has both an
`is_<system>()` predicate (returns logical) and an
`ensure_is_<system>()` guard (errors otherwise):

```{r}
is_polar(pol)
is_cartesian_2d(cart)
```

## Where to next?

- [The metadata attribute](aniframe-metadata.html) — units, origin,
  sampling rate, and the setters that keep them consistent with the
  data.
- [Connections](aniframe-connections.html) — recording skeleton edges
  or other relationships between values of an identity / temporal
  variable.
