---
title: "Metadata on an aniframe"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aniframe)
```

## Why an attribute, not columns?

Every `aniframe` carries a metadata list as an R attribute alongside
the data columns. The metadata records the things that are true of the
recording as a whole rather than of any single observation: the source
software, the sampling rate, what units the spatial coordinates are
in, where the coordinate origin sits, and so on.

Keeping this information attached to the object — rather than living
in a separate file or being passed around as extra arguments — is what
lets the rest of the _animovement_ ecosystem stay loosely coupled.
A reader (in _aniread_) populates the metadata at load time, and any
downstream tool can read it back without a hand-off.

This article covers the metadata attribute and the functions that read
and update it. For the data-column structure see
[The aniframe data structure](aniframe_structure.html); for the
`connections` field specifically see
[Connections](aniframe_connections.html).

## The metadata attribute

You can see the full metadata by printing it directly:

```{r}
data <- example_aniframe()
get_metadata(data)
```

The fields and their defaults are defined in one place,
`default_metadata()` — that's the canonical source of truth for what
an `aniframe`'s metadata looks like.

```{r}
str(default_metadata(), max.level = 1)
```

The fields fall into a few groups:

| Group              | Fields                                                              |
|--------------------|---------------------------------------------------------------------|
| **Provenance**     | `source`, `source_version`, `filename`, `start_datetime`            |
| **Sampling**       | `sampling_rate`                                                     |
| **Units**          | `unit_space`, `unit_time`, `unit_angle`                             |
| **Frame of reference** | `reference_frame`, `coordinate_system`, `origin`, `y_height`    |
| **Slot vocabulary**| `variables_what`, `variables_when`, `variables_where`               |
| **Relationships**  | `connections`                                                       |

`filename` accepts a character vector — readers like
`aniread::read_trackball()` populate it with all source paths.

## Reading and writing metadata

`get_metadata()` and `set_metadata()` are the workhorses.

```{r}
get_metadata(data, "sampling_rate")

data <- set_metadata(data, sampling_rate = 30, source = "deeplabcut")
get_metadata(data, "sampling_rate")
get_metadata(data, "source")
```

`set_metadata()` validates the input — factor fields are checked
against their permitted levels, and unknown fields are rejected.

For fields whose update has side effects on the data columns (or on
related fields), prefer the dedicated setters listed below.

| Setter                | Touches                                                        |
|-----------------------|----------------------------------------------------------------|
| `set_unit_space()`    | converts `x`/`y`/`z` between length units                      |
| `set_unit_time()`     | converts `time` between time units                             |
| `set_unit_angle()`    | converts `phi`/`theta` (auto) and any extra `cols` you supply  |
| `set_sampling_rate()` | flips `unit_time` from frames to seconds and rescales `time`   |
| `set_origin()`        | flips the y-axis around `y_height` when changing convention    |
| `set_y_height()`      | sets the recorded frame height used by `set_origin()`          |

## Coordinate origin

For 2D image-derived data there's an annoying convention split: most
image / video tooling uses the **top-left** corner as `(0, 0)` (y
increases downward), while plotting and most maths uses the
**bottom-left** corner (y increases upward). `aniframe` records which
one your data uses in the `origin` field, with permitted values
`c("bottom_left", "top_left")`.

`set_origin()` does the actual flip when you change convention. It
needs the frame height to compute `y_new = y_height - y_old`, so
`y_height` must be set first. Readers populate it automatically; for
manually-constructed `aniframe` objects, `as_aniframe()` falls back to
`max(y)`, and `set_y_height()` lets you override that with the true
value.

```{r}
img <- aniframe(
  individual = 1L, time = 1:4,
  x = c(0, 10, 20, 30),
  y = c(50, 100, 150, 200)
) |>
  set_y_height(1080)

img$y
img <- set_origin(img, "top_left")
img$y # reflected: 1080 - original y
```

## Units

Spatial, temporal, and angular units each have their own setter.
Conversions between standard units are automatic; conversions from
unknown / `frame` / `px` units require an explicit calibration factor
(or, for time, a `sampling_rate`).

```{r}
data <- example_aniframe(n_dims = 2) |>
  set_metadata(unit_space = "mm")

data_cm <- set_unit_space(data, to_unit = "cm")
get_metadata(data_cm, "unit_space")
```

```{r}
data <- example_aniframe() # default unit_time = "frame"
data_s <- set_sampling_rate(data, sampling_rate = 30)
get_metadata(data_s, "unit_time") # now "s"
range(data_s$time) # frames divided by fps
```

Spatial angular columns (`phi`, `theta`) are converted automatically
by `set_unit_angle()` whenever they're present. Pass `cols` only for
non-spatial angular columns (e.g. heading direction).

```{r}
pol <- aniframe(
  individual = 1L, time = 1:3,
  rho = c(1, 1, 1), phi = c(0, pi / 2, pi)
)
pol$phi
pol_deg <- set_unit_angle(pol, to_unit = "deg")
pol_deg$phi
```
